---
title: "Homework 4: Diffusion of Tetracycline"
author: Yakitori
output: pdf_document
---

We continue examining the diffusion of tetracycline among doctors in Illinois in the early 1950s, building on our work in lab 6.  You will need the data sets `ckm_nodes.csv` and `ckm_network.dat` from the labs.

<br/>

<br/>

1. Clean the data to eliminate doctors for whom we have no adoption-date information, as in the labs.  Only use this cleaned data in the rest of the assignment.

<br/>

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
# read in the data
ckm_nodes <- read.csv('data/ckm_nodes.csv', stringsAsFactors = T)
ckm_network <- read.table('data/ckm_network.dat')
# clean the data
index <- which(!is.na(ckm_nodes$adoption_date))
ckm_network <- ckm_network[index, index]
ckm_nodes <- ckm_nodes %>% filter(!is.na(adoption_date))
```

<br/>

2. Create a new data frame which records, for every doctor, for every month, whether that doctor began prescribing tetracycline that month, whether they had adopted tetracycline before that month, the number of their contacts who began prescribing strictly _before_ that month, and the number of their contacts who began prescribing in that month or earlier.  Explain why the dataframe should have 6 columns, and 2125 rows.  Try not to use any loops.

<br/>

```{r}
# initialization
ckm_detail <- matrix(NA, ncol = 2, nrow = 2125) %>% as.data.frame
colnames(ckm_detail) <- c('doctor', 'month')
ckm_detail$doctor <- rep(1:125, each = 17)
ckm_detail$month <- rep(1:17, times = 125)

# whether that doctor began prescribing tetracycline that month
# and whether they had adopted tetracycline before that month
ckm_detail <- ckm_detail %>% 
    mutate(adoption_date = rep(ckm_nodes$adoption_date, each = 17)) %>%
    mutate(adoption_now = adoption_date <= month) %>% 
    mutate(adoption_before = adoption_date < month) %>% 
    select(-adoption_date)
```

```{r}
# This function is to compute the prescribing information of a doctor's contacts
# index is the number of th doctor
contact_adoption <- function(index, column_name = ''){
  # Exception handling
  if(! column_name %in% colnames(ckm_detail)){
    print('Column names error!')
    return()
  }
  # extract data
  adoption_index <- which(ckm_detail$doctor %in% which(ckm_network[index, ] == 1))
  contact_info <- ckm_detail[column_name][adoption_index, 1] %>% 
    matrix(nrow = 17) %>% rowSums
  return(contact_info)
}
```

```{r}
# the number of their contacts who began prescribing strictly before that month
# the number of their contacts who began prescribing in that month or earlier
contact_adoption_now <- sapply(1:125, contact_adoption, 'adoption_now') %>% 
  matrix(ncol = 1)
contact_adoption_before <- sapply(1:125, contact_adoption, 'adoption_before') %>% 
  matrix(ncol = 1)
ckm_detail$contact_adoption_now <- contact_adoption_now
ckm_detail$contact_adoption_before <- contact_adoption_before

str(ckm_detail)
```

<br/>

The dataframe ckm_detail should have 6 columns, which is for the doctor number, the month, whether that doctor began prescribing tetracycline that month, whether they had adopted tetracycline before that month, the number of their contacts who began prescribing in that month or earlier the number of their contacts who began prescribing strictly _before_ that month.

Also it should have 2125 rows since there are 125 $\times$ 17 observations for 125 doctors and 17 months.
<br/>

<br/>


3.  Let
    \[
    p_k = \Pr(\text{A doctor starts prescribing tetracycline this month} \mid \\
    \text{Number of doctor's contacts prescribing before this month}=k)
    \]
    and
   \[
     q_k = \Pr(\text{A doctor starts prescribing tetracycline this month} \mid \\ \text{Number of doctor's contacts prescribing this month}=k)
    \]
   We suppose that $p_k$ and $q_k$ are the same for all months.
<br/>

     a. Explain why there should be no more than $21$ values of $k$ for which we can estimate $p_k$ and $q_k$ directly from the data.
     
```{r}
apply(ckm_network, 1, sum) %>% max
```
A doctor has $20$ contacts at most, hence there should be no more than $21$ values of $k$ from $0$ to $20$.
<br/>

    
     b. Create a vector of estimated $p_k$ probabilities, using the data frame from (2).  Plot the probabilities against the number of prior-adoptee contacts $k$.
<br/>

```{r}
# This function is used to compute the probability p_k or q_k
# Also to tackle the following problem, I add a function block to compute n_k
compute_pk_or_qk <- function(k, colname = '', problem_type = 1){
    df <- ckm_detail$adoption_now[which(ckm_detail[colname] == k)]
    if(problem_type == 1) return(sum(df) / length(df))
    # the else prat is used to tackle the subsequent problem
    # It is used to compute n_k
    else return(ckm_detail$doctor[which(ckm_detail[colname] == k)] %>% unique %>% length)
}
```

```{r}
# compute the probability
ckm_contact <- data.frame(k = 0:20, pk = rep(NA, 21), qk = rep(NA, 21))
ckm_contact$pk <- sapply(0:20, compute_pk_or_qk, 'contact_adoption_now', 1)
# plot
ckm_contact %>% ggplot() + 
    geom_point(aes(x = k, y = pk)) + 
    labs(x = 'Number of doctor\'s contacts prescribing before this month', y = 'pk')
```

<br/>
     
     c. Create a vector of estimated $q_k$ probabilities, using the data frame from (2). Plot the probabilities against the number of prior-or-contemporary-adoptee contacts $k$.

<br/>

```{r}
# compute qk
ckm_contact$qk <- sapply(0:20, compute_pk_or_qk, 'contact_adoption_before', 1)
# plot
ckm_contact %>% ggplot() + 
    geom_point(aes(x = k, y = qk)) + 
    labs(x = 'Number of doctor\'s contacts prescribing before this month', y = 'qk')
```

<br/>

<br/>

4. Because it only conditions on information from the previous month, $p_k$ is a little easier to interpret than $q_k$.  It is the probability per month that a doctor adopts tetracycline, if they have exactly $k$ contacts who had already adopted tetracycline.
<br/>


    a. Suppose $p_k = a + bk$.  This would mean that each friend who adopts the new drug increases the probability of adoption by an equal amount.  Estimate this model by least squares, using the values you constructed in (3b).  Report the parameter estimates.
<br/>

```{r}
ols.fit <- lm(pk ~ k, data = ckm_contact)
summary(ols.fit)
```
<br/>
    
    b. Suppose $p_k = e^{a+bk}/(1+e^{a+bk})$.  Explain, in words, what this model would imply about the impact of adding one more adoptee friend on a given doctor's probability of adoption.  (You can suppose that $b > 0$, if that makes it easier.)  Estimate the model by least squares, using the values you constructed in (3b).
<br/>
    
```{r}
logit.fit <- glm(pk ~ k, family = binomial(link = 'logit'), data = ckm_contact)
summary(logit.fit)
```
<br/>

The coefficient of $k$ is positive, which means adding one more adoptee friend wolud increase the given doctor's probability of adoption.
<br/>

    c. Plot the values from (3b) along with the estimated curves from (4a) and (4b).  (You should have one plot, with $k$ on the horizontal axis, and probabilities on the vertical axis.)  Which model do you prefer, and why?
<br/>

```{r}
# OLS prediction and Logistic prediction
ols.pred <- predict(ols.fit)
logit.pred <- predict(logit.fit, type = 'response')
ckm_pred <- data.frame(k = 0:17, real_pk = na.omit(ckm_contact$pk), 
                       ols_pred = ols.pred, logit_pred = logit.pred)
# plot
ckm_pred %>% ggplot() + geom_point(aes(x = k, y = real_pk)) + 
    geom_line(aes(x = k, y = ols_pred, colour = 'red'), size = 1.1) + 
    geom_line(aes(x = k, y = logit_pred, colour = 'blue'), size = 1.1) + 
    labs(y = 'pk', title = 'Predicting pk by Linear Regression and Logistic Regression') + 
    scale_color_manual(name = "Model", values = c('red' = 'red', 'blue' = 'blue'), 
                       labels = c('red'='Linear Regression', 'blue'='Logistic Regression'))

```
<br/>

Due to the missing value or Inf value, some data is removed in the plot. 

I prefer logistic regression since as a classification model, it can better fit the data. Also its value ranges form $0$ to $1$, which is more comprehensible and precise than that of linear regression.

<br/>

<br/>


_For quibblers, pedants, and idle hands itching for work to do_: The $p_k$ values from problem 3 aren't all equally precise, because they come from different numbers of observations.  Also, if each doctor with $k$ adoptee contacts is independently deciding whether or not to adopt with probability $p_k$, then the variance in the number of adoptees will depend on $p_k$.  Say that the actual proportion who decide to adopt is $\hat{p}_k$.  A little probability (exercise!) shows that in this situation, $\mathbb{E}[\hat{p}_k] = p_k$, but that $\mathrm{Var}[\hat{p}_k] = p_k(1-p_k)/n_k$, where $n_k$ is the number of doctors in that situation.  (We estimate probabilities more precisely when they're really extreme [close to 0 or 1], and/or we have lots of observations.)  We can estimate that variance as $\hat{V}_k = \hat{p}_k(1-\hat{p}_k)/n_k$.  Find the $\hat{V}_k$, and then re-do the estimation in (4a) and (4b) where the squared error for $p_k$ is divided by $\hat{V}_k$.  How much do the parameter estimates change?  How much do the plotted curves in (4c) change?

<br/>

```{r}
# compute the variance
ckm_contact <- ckm_contact %>% 
  mutate(nk = sapply(0:20, compute_pk_or_qk, 'contact_adoption_now', 2)) %>% 
  mutate(Vk = pk * (1 - pk) / nk)
head(ckm_contact)
```

<br/>

We can transform the question equivalently by replicating each $(k, \ p_k)$ for $n_k$ times respectively.

<br/>

```{r}
# adjust the sample according to the number of doctors
numbers_adjusted <- function(vec, times){
  result <- c()
  for(i in 1:length(times)){
    result <- c(result, rep(vec[i], times[i]))
  }
  return(result)
}
```

```{r}
# new data frame
ckm_contact_2 <- data.frame(k = numbers_adjusted(ckm_contact$k, ckm_contact$nk), 
                            pk = numbers_adjusted(ckm_contact$pk, ckm_contact$nk))
str(ckm_contact_2)
```

<br/>

```{r}
# linear regression
ols.fit <- lm(pk ~ k, data = ckm_contact_2)
summary(ols.fit)
# logistic regression
logit.fit <- glm(pk ~ k, family = binomial(link = 'logit'), data = ckm_contact_2)
summary(logit.fit)
```

<br/>

The parameters of linear regression model keep almost unchanged, while the coefficients of
logistic regression become much more significant, implying a better-fitted model than before.
<br/>

```{r}
# OLS prediction and Logistic prediction
ols.pred <- predict(ols.fit)
logit.pred <- predict(logit.fit, type = 'response')
ckm_pred <- data.frame(k = na.omit(ckm_contact_2$k), real_pk = na.omit(ckm_contact_2$pk), 
                       ols_pred = ols.pred, logit_pred = logit.pred)
# plot
ckm_pred %>% ggplot() + geom_point(aes(x = k, y = real_pk)) + 
    geom_line(aes(x = k, y = ols_pred, colour = 'red'), size = 1.1) + 
    geom_line(aes(x = k, y = logit_pred, colour = 'blue'), size = 1.1) + 
    labs(y = 'pk', title = 'Predicting pk by Linear Regression and Logistic Regression') + 
    scale_color_manual(name = "Model", values = c('red' = 'red', 'blue' = 'blue'), 
                       labels = c('red'='Linear Regression', 'blue'='Logistic Regression'))

```

<br/>

The graph above can verify what we comment on the model summary. It is obvious that compared to 4(c), logistic regression fits the data better, but linear regression become invalid after $k = 10$.
